Unbiased pangenome graphs

### 2022-11-14 manuscript/main.tex
get it going
remove links from the references

\section{Introduction}
\label{sec:introduction}
A pangenome models the full genomic information of a species or clade \citep{Medini_2005,Sherman_2020}.
In contrast to reference-based approaches that relate sequences to a particular reference genome, methods that use pangenome reference systems attempt to model the mutual relationship between all represented genomes \citep{cpang2018}.
%These models encode the mutual relationships between all the genomes represented, in contrast to reference-based approaches which relate sequences only to a particular ``reference'' genome.
%In principle, pangenome graphs can allow for a great reduction in reference bias, allowing any part of a pangenome to be considered in biological studies. %, and may provide a simpler representation of structurally-variable regions.
Many approaches model the pangenome alignment as a \textit{pangenome graph} \citep{Garrison_2018,Yokoyama2019,Hickey:2020}.
A pangenome graph encodes DNA sequences as walks through an underlying language encoded in a sequence graph \citep{Hein_1989}.
In a pangenome graph, variation can be understood in the context of any part of any included genome  \citep{Eizenga_2020}.
This lets us avoid the problem of reference bias, which can be understood as the limitation of analyses to genome sequences that are similar to a chosen reference genome.
%Reference bias distorts an inferred genome to appear more similar to the reference genome than it actually is.

@@

%A practical implementation introduces a new set of problems related to input and output. % that we have not yet introduced.
Our implementation reads standard data formats, FASTA or FASTQ for the input sequences, and PAF \citep{Li_2018} for pairwise alignments.
It writes the graph in standard Graphical Fragment Assembly (GFA) format \citep{GFA}.

@@

In our implementation, we rely on several basic external memory kernels.
To reduce working memory requirements to an absolute minimum, we use a disk-backed version of the implicit interval tree that memory-maps the sorted array of intervals \citep{mmmulti}.
Indexing the implicit interval tree requires a sorting step which dominates the runtime of our algorithm.
We adapt the current best-performing in-place parallel sorting algorithm, In-place Parallel Super Scalar Samplesort (IPS$^4$o), to work on a disk-backed, memory-mapped array \citep{axtmann2017}.
This allows us work with $\mathcal{A}$, $\mathcal{Z}$, and $\bar{\mathcal{Z}}$ in external memory.
By storing pairs of numerical identifiers in the backing array, we are able to generate a disk-backed multiset model which we use to compute the unique set of edges $\mathcal{E}$ in terms of offsets in $\mathcal{V}$.
The graph sequence vector $\mathcal{V}$ is simply written by appending characters to a file.
We mark nodes to generate $\mathcal{V}'$ using a bitvector kept in main memory, over which we subsequently generate a rank/select dictionary \citep{Gog_2014} for support of the final emission of the graph $\mathcal{G}$.

@@

We mitigate this issue with a simple filter, \textit{seqwish -k}, which simply ignores exact matches that are shorter than $k$ characters.
This filter necessarily increases the size of the induced graph.
But, it also replaces complex motifs shorter than $k$ with single bubbles.
In doing so, it also removes short, expensive matches, reducing the overall space requirements for \textit{seqwish}.
When set very high, this filter can be used to generate a coarse, high-confidence graph built only from very long exact matches which will tend to be unique in the genome.
Although the application of the $k>$ filter can result in a graph that is relatively ``under-aligned'', we can further refine it through the application of local multiple sequence alignment \citep{Gao_2020}, or graph normalization \citep{Doerr_2021}.
In a pangenomic context, underalignment caused by $k>$ match filtering can be mitigated by transitive relationships present in the pangenome.

@@

%    \FIXME{To integrate to the text}
%    To support this kind of data scale, we implement \texttt{seqwish}'s algorithm using disk-backed memory.
%    We write the alignment graph into a disk-backed implicit interval tree, wherein each match in the input alignments is a queryable range.
%    A lock-free, parallel union-find algorithm \FIXME{Wait-free parallel algorithms for the union-find problem} allows us to, for each base pair of input sequence, collect all other base pairs in the pangenome that are aligned to it.
%    By processing this transitive closure operation in chunks (e.g. 10Mbp at a time), we limit the maximum working memory (RAM) requirements of the process without affecting the output.
%    After the transitive closure, we write our final graph on disk in a set of arrays, then augment it with the paths of our input sequences,
%    finally emitting the resulting graph in the standard Graphical Fragment Assembly (GFA) format~\citep{GFA}.

@@

To construct the graph we first generate alignments with \textit{wfmash} \citep{wfmash}, a DNA sequence aligner designed specifically for high performance all-to-all alignment of fully-assembled genomes.
\textit{wfmash} combines an algorithm for generating whole-genome homology maps (\textit{MashMap2}) \citep{Jain_2018} with an extension of the wavefront algorithm (WFA) \citep{Marco_Sola_2020} capable of obtaining base-level alignments for whole chromosomes.
\textit{MashMap2} allows the user to define a homology length and pairwise divergence, expressed as a percent identity, over which to generate homology maps.
This is useful when constructing pangenome graphs, because, in contrast to methods that are based on $k$-mer chaining \citep{harris2007lastz,Li_2018}, it allows us to query the homology space of input genomes using two easily-interpretable parameters.
The version of \textit{wfmash} used in these experiments allows us to align sequences with up to 10\% divergence between them, providing highly-sensitive input for our experiments.
Our use of \textit{wfmash} is pragmatic, and any mapping method capable of generating PAF with base-level alignments in CIGAR strings is capable of being used as input to \textit{seqwish}.

@@

To focus on these questions, we examine the case of a single chromosome (chrV) of \textit{S. cerevisiae} for which we have 7 assemblies \citep{Yue_2017}.
We compute 100 random permutations \citep{Williams2009-ou} of these assemblies and provide them to several graph construction methods, including \textit{minigraph} \citep{Li:2020}, \textit{TwoPaCo} \citep{Minkin_2016}, and \textit{seqwish} based on \textit{minimap2} \citep{Li_2018} or \textit{wfmash} \citep{wfmash} alignments.

>>>>

\section{Introduction}
\label{sec:introduction}
A pangenome models the full genomic information of a species or clade \citep{Medini_2005,Sherman_2020,Garrison_2019_thesis}.
In contrast to reference-based approaches that relate sequences to a particular reference genome, methods that use pangenome reference systems attempt to model the mutual relationship between all represented genomes \citep{cpang2018}.
%These models encode the mutual relationships between all the genomes represented, in contrast to reference-based approaches which relate sequences only to a particular ``reference'' genome.
%In principle, pangenome graphs can allow for a great reduction in reference bias, allowing any part of a pangenome to be considered in biological studies. %, and may provide a simpler representation of structurally-variable regions.
Many approaches model the pangenome alignment as a \textit{pangenome graph} \citep{Garrison_2018,Yokoyama2019,Hickey:2020}.
A pangenome graph encodes DNA sequences as walks through an underlying language encoded in a sequence graph \citep{Hein_1989}.
In a pangenome graph, variation can be understood in the context of any part of any included genome  \citep{Eizenga_2020}.
This lets us avoid the problem of reference bias, which can be understood as the limitation of analyses to genome sequences that are similar to a chosen reference genome.
%Reference bias distorts an inferred genome to appear more similar to the reference genome than it actually is.

@@

%A practical implementation introduces a new set of problems related to input and output. % that we have not yet introduced.
Our implementation reads standard data formats, FASTA or FASTQ for the input sequences, and PAF \citep{Li_2018} for pairwise alignments.
It writes the graph in standard Graphical Fragment Assembly (GFA) format (\url{https://github.com/GFA-spec/GFA-spec}).

@@

In our implementation, we rely on several basic external memory kernels.
To reduce working memory requirements to an absolute minimum, we use a disk-backed version of the implicit interval tree that memory-maps the sorted array of intervals (\url{https://github.com/ekg/mmmulti}).
Indexing the implicit interval tree requires a sorting step which dominates the runtime of our algorithm.
We adapt the current best-performing in-place parallel sorting algorithm, In-place Parallel Super Scalar Samplesort (IPS$^4$o), to work on a disk-backed, memory-mapped array \citep{axtmann2017}.
This allows us work with $\mathcal{A}$, $\mathcal{Z}$, and $\bar{\mathcal{Z}}$ in external memory.
By storing pairs of numerical identifiers in the backing array, we are able to generate a disk-backed multiset model which we use to compute the unique set of edges $\mathcal{E}$ in terms of offsets in $\mathcal{V}$.
The graph sequence vector $\mathcal{V}$ is simply written by appending characters to a file.
We mark nodes to generate $\mathcal{V}'$ using a bitvector kept in main memory, over which we subsequently generate a rank/select dictionary \citep{Gog_2014} for support of the final emission of the graph $\mathcal{G}$.

@@

We mitigate this issue with a simple filter, \textit{seqwish -k}, which simply ignores exact matches that are shorter than $k$ characters.
This filter necessarily increases the size of the induced graph.
But, it also replaces complex motifs shorter than $k$ with single bubbles.
In doing so, it also removes short, expensive matches, reducing the overall space requirements for \textit{seqwish}.
When set very high, this filter can be used to generate a coarse, high-confidence graph built only from very long exact matches which will tend to be unique in the genome.
Although the application of the $k>$ filter can result in a graph that is relatively ``under-aligned'', we can further refine it through the application of local multiple sequence alignment \citep{Gao_2020}, or graph normalization (\url{https://github.com/marschall-lab/GFAffix}).
In a pangenomic context, underalignment caused by $k>$ match filtering can be mitigated by transitive relationships present in the pangenome.

@@

%    \FIXME{To integrate to the text}
%    To support this kind of data scale, we implement \texttt{seqwish}'s algorithm using disk-backed memory.
%    We write the alignment graph into a disk-backed implicit interval tree, wherein each match in the input alignments is a queryable range.
%    A lock-free, parallel union-find algorithm \FIXME{Wait-free parallel algorithms for the union-find problem} allows us to, for each base pair of input sequence, collect all other base pairs in the pangenome that are aligned to it.
%    By processing this transitive closure operation in chunks (e.g. 10Mbp at a time), we limit the maximum working memory (RAM) requirements of the process without affecting the output.
%    After the transitive closure, we write our final graph on disk in a set of arrays, then augment it with the paths of our input sequences,
%    finally emitting the resulting graph in the standard Graphical Fragment Assembly (GFA) format (\url{https://github.com/GFA-spec/GFA-spec}).

@@

To construct the graph we first generate alignments with \textit{wfmash} (\url{https://github.com/waveygang/wfmash}), a DNA sequence aligner designed specifically for high performance all-to-all alignment of fully-assembled genomes.
\textit{wfmash} combines an algorithm for generating whole-genome homology maps (\textit{MashMap2}) \citep{Jain_2018} with an extension of the wavefront algorithm (WFA) \citep{Marco_Sola_2020} capable of obtaining base-level alignments for whole chromosomes.
\textit{MashMap2} allows the user to define a homology length and pairwise divergence, expressed as a percent identity, over which to generate homology maps.
This is useful when constructing pangenome graphs, because, in contrast to methods that are based on $k$-mer chaining \citep{harris2007lastz,Li_2018}, it allows us to query the homology space of input genomes using two easily-interpretable parameters.
The version of \textit{wfmash} used in these experiments allows us to align sequences with up to 10\% divergence between them, providing highly-sensitive input for our experiments.
Our use of \textit{wfmash} is pragmatic, and any mapping method capable of generating PAF with base-level alignments in CIGAR strings is capable of being used as input to \textit{seqwish}.

@@

To focus on these questions, we examine the case of a single chromosome (chrV) of \textit{S. cerevisiae} for which we have 7 assemblies \citep{Yue_2017}.
We compute 100 random permutations \citep{Williams2009-ou} of these assemblies and provide them to several graph construction methods, including \textit{minigraph} \citep{Li:2020}, \textit{TwoPaCo} \citep{Minkin_2016}, and \textit{seqwish} based on \textit{minimap2} \citep{Li_2018} or \textit{wfmash} (\url{https://github.com/waveygang/wfmash}) alignments.
